---
title: "Ectoderm trajectory"
---
  
```{r load_modules, echo=FALSE, include=FALSE}
library(SingleCellExperiment)
library(scran)
library(scater)
library(destiny)
```

# Define settings

Define I/O
```{r define_io, echo=FALSE}
source("/Users/ricard/gastrulation_multiome_10x/settings.R")
io$outdir <- paste0(io$basedir,"/results/rna/trajectories/ectoderm_trajectory")
```

Define options
```{r}
# opts$celltypes = c(
#   "Epiblast",
#   "Rostral_neurectoderm",
#   # "Neural_crest",
#   "Forebrain_Midbrain_Hindbrain"
#   # "Spinal_cord",
#   # "Surface_ectoderm"
# )

opts$stage_celltype = c(
  "E7.5_Epiblast",
  "E7.5_Rostral_neurectoderm",
  "E8.5_Rostral_neurectoderm",
  "E8.5_Forebrain_Midbrain_Hindbrain"
)

opts$stage_celltype.colors <- c(
  "E7.5_Epiblast" = "#635547",
  "E7.5_Rostral_neurectoderm" = "#65A83E",
  "E8.5_Rostral_neurectoderm" = "#4a941f",
  "E8.5_Forebrain_Midbrain_Hindbrain" = "#647a4f"
)

opts$batch_correction <- FALSE
```

Update sample metadata
```{r load_metadata, echo=FALSE}
sample_metadata <- fread(io$metadata) %>%
  .[pass_rnaQC==TRUE & doublet_call==FALSE] %>%
  .[,stage_celltype:=paste(stage,celltype.mapped,sep="_")] %>%
  # .[celltype.mapped%in%opts$celltypes & sample%in%opts$samples]
  .[stage_celltype%in%opts$stage_celltype] %>%
  .[,stage_celltype:=factor(stage_celltype,levels=opts$stage_celltype)]
table(sample_metadata$stage_celltype)
```

# Load data

Load RNA expression data as SingleCellExperiment object
```{r load_data, echo=FALSE}
sce <- load_SingleCellExperiment(
  file = io$sce, 
  normalise = TRUE, 
  cells = sample_metadata$cell, 
  remove_non_expressed_genes = TRUE
)
dim(sce)
```

Add sample metadata to the colData of the SingleCellExperiment
```{r}
colData(sce) <- sample_metadata %>% as.data.frame %>% tibble::column_to_rownames("cell") %>%
  .[colnames(sce),] %>% DataFrame()
```

# Parse data

Select HVG
```{r}
decomp <- modelGeneVar(sce)
decomp <- decomp[decomp$mean > 0.01,]
hvgs <- rownames(decomp)[decomp$p.value <= 0.05]

# Subset SingleCellExperiment
sce_filt <- sce[hvgs,]
dim(sce_filt)
```

Save HVGs
```{r}
hvgs.to.save <- rownames(decomp)[decomp$p.value <= 0.01]
saveRDS(hvgs.to.save, paste0(io$outdir,"/hvgs.rds"))
```

Create data.table for downstream analysis
```{r}
rna_dt <- as.matrix(logcounts(sce_filt)) %>% t %>% 
  # scale(center = FALSE, scale = TRUE) %>%
  as.data.table(keep.rownames="cell") %>% 
  melt(id.vars="cell", value.name="expr", variable.name="gene")
```


# PCA

```{r}
if (opts$batch_correction) {
  if (length(unique(sce_filt$sample))>1) {
    library(batchelor)
    pca <- multiBatchPCA(sce_filt, batch=sce_filt$sample)
    pca.corrected <- reducedMNN(pca)$corrected
    colnames(pca.corrected) <- paste0("PC",1:ncol(pca.corrected))
    reducedDim(sce_filt, "PCA") <- pca.corrected
  }
} else {
  sce_filt <- runPCA(sce_filt, ncomponents = 5, ntop=nrow(sce_filt))
}
```

Plot PCA
```{r message=FALSE}
p <- plotPCA(sce_filt, colour_by="celltype.mapped", ncomponents = c(1,2)) +
  scale_colour_manual(values=opts$celltype.colors)

pdf(sprintf("%s/pca.pdf",io$outdir), width=8, height=5)
print(p)
dev.off()
```

# UMAP

Run UMAP
```{r}
set.seed(42)
sce_filt <- runUMAP(sce_filt, dimred="PCA", n_neighbors = 25, min_dist = 0.35)
```

Plot UMAP
```{r message=FALSE}
p <- plotUMAP(sce_filt, colour_by="celltype.mapped", ncomponents = c(1,2)) +
  scale_colour_manual(values=opts$celltype.colors)

pdf(sprintf("%s/umap.pdf",io$outdir), width=8, height=5)
print(p)
dev.off()
```

# Diffusion pseudotime

Calculate Diffusion map
```{r}
set.seed(42)
dm <- DiffusionMap(sce_filt, n_pcs=2)
```

Add to the SingleCellExperiment object
```{r}
reducedDim(sce_filt, "DiffusionMap") <- dm@eigenvectors[,c(1,2)]
```

Plot Diffusion map
```{r message=FALSE}
p <- plotReducedDim(sce_filt, dimred = "DiffusionMap", colour_by="celltype.mapped", ncomponents = c(1,2)) +
  scale_colour_manual(values=opts$celltype.colors)
# plotReducedDim(sce_filt, dimred = "DiffusionMap", colour_by="stage_celltype", ncomponents = c(1,2)) +
#   scale_colour_manual(values=opts$stage_celltype.colors)

pdf(sprintf("%s/diffusion_map.pdf",io$outdir), width=8, height=5)
print(p)
dev.off()
```

Calculate ranks
```{r}
pseudotime.dt <- data.table(
  cell = colnames(sce_filt),
  V1 = -eigenvectors(dm)[,1]
) %>% setorder(V1) %>% .[,rank_V1:=1:.N]
```

Load precomputed pseudotime
```{r}
# pseudotime.dt <- fread(paste0(io$outdir,"/ectoderm_trajectory.txt.gz"))
```

Boxplots of pseudotime values per celltype

```{r}
to.plot1 <- pseudotime.dt %>%
  .[,c("cell","V1","rank_V1")] %>% 
  merge(sample_metadata[,c("cell","celltype.mapped","stage")], by="cell")

p <- ggboxplot(to.plot1, x="celltype.mapped", y="V1", fill="celltype.mapped", outlier.shape=NA) +
  facet_wrap(~stage, nrow=1, scales="free_x") +
  scale_fill_manual(values=opts$celltype.colors) +
  labs(x="", y="") +
  theme_classic() +
  guides(x = guide_axis(angle = 90)) +
  theme(
    legend.position = "right",
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank()
  )

pdf(sprintf("%s/pseudotime_boxplot_celltype.pdf",io$outdir), width=8, height=5)
print(p)
dev.off()
```

Scatterplots of pseudotime values versus expression of marker genes

```{r}
diff.dt <- fread("/Users/ricard/data/gastrulation10x/results/differential/celltypes/all_stages/Epiblast_vs_Forebrain_Midbrain_Hindbrain.txt.gz")

genes.to.plot <- diff.dt %>%
  .[abs(logFC)>2 & sig==T,gene] %>%
  head(n=28)

to.plot2 <- to.plot1 %>% merge(rna_dt[gene%in%genes.to.plot],by="cell")
```


```{r}
p <- ggplot(to.plot2, aes(x=V1, y=expr)) +
  geom_point(aes(fill=expr), size=1, shape=21, stroke=0.1, alpha=0.5) +
  stat_smooth(aes(fill=expr), method="loess", color="black", alpha=0.90, span=0.5) +
  geom_rug(aes(color=celltype.mapped), sides="b") +
  viridis::scale_fill_viridis() +
  scale_color_manual(values=opts$celltype.colors) +
  facet_wrap(~gene, scales="free_y") +
  labs(x="Pseudotime", y="Gene expression") +
  theme_classic() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position="none"
  )

pdf(sprintf("%s/pseudotime_vs_expression.pdf",io$outdir), width=10, height=8)
print(p)
dev.off()
```

Save coordinates
```{r}
fwrite(pseudotime.dt, paste0(io$outdir,"/ectoderm_trajectory.txt.gz"))
```

